Relationships in the extremes and their influence on competition and coexistence

### 2021-08-04 paper/Paper.Rnw
small

%Put other possible titles below, commented, and we will select the best one at the end
%Relationships between extreme events and their influence on competition and coexistence
%Coexistence and consequences of extreme environmental relationships
%Coexistence and consequences of environmental relationships in their extremes
%Coexistence and implications of asymmetric environmental relationships
%Extreme [climatic] events and their implications for biodiversity
%Deductions to coexistence, extreme events, and asymmetric relationships
%Extreme relationships (relationships in their extremes) and their deductions ([negative] contributions) to coexistence

@@

%Old version, prior to my edits: Central to ecology is the maintenance of biodiversity in competitive ecosystems. Populations subject to strong
%competitive pressure may succumb to competitive exclusion - an idea which postulates if species within the same
%trophic level and similar niches exist, one will eventually outcompete the other (Hardin, 1960). Mediation of
%competitive pressure and exclusion allow for species to stably coexist. The criteria for coexistence and competitive
%exclusion can be broken down into contributing mechanisms provided by modern coexistence theory, a framework
%which seeks to assess and quantify mechanisms that result in coexisting species and thus the maintenance of
%biodiversity (Chesson, 1994).
%Now the new, edited version, with some additional changes requested, to be made by Jasmin, please:
The maintenance of biodiversity in competitive ecosystems is a research topic at the absolute core of modern ecology (\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
%***DAN: Jasmin, add a few classic references. See the first sentence of Ellner et al 2019, read the abstracts of those papers
%(more than the abstracts if you want), and consider using them as references here.
The competitive exclusion principle, covered, for instance, in most or all introductory ecology courses and texts, states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will outcompete 
the other and drive it extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how stable species coexistence 
nevertheless occurs, and biodiversity in maintained, in real ecosystems, through stabilizing mechanisms and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with name such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to quite a few empirical systems, including annual plants \citep{Angert_2009, Hallett_2019}, perennial plants \citep{Adler_2010, Adler_2006}, tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{}. 
%***DAN: Jasmin, fill in Y and Y with names of systems, and accompanying reference(s), 
%in a similar way to how I filled in the Angert example.
%Please also read the abstract of the Angert reference to make sure it is appropriate to cite it in this way (and so that
%you become aware of that reference). One place to look for more references to put in for Y and Z is the top of the 
%right column on the first page of the Barabas review (Ecol Monogr, 2018).
even though it is mathematically complex in the eyes
of most ecologists (but see \cite{Ellner_2016,Ellner_2019} for efforts to make it more readily applied).
%***DAN: Jasmin, note the differences carefully, please, between your version (commented, above) and the edits I've 
%made. It's the same in main structure, making all the same points. It's just a bit more fleshed out. One of the purposes
%of an Intro is to link the reader to the appropriate background lit. You need a fair number of references and a little
%more detail than you previously gave to do that.

@@

%***DAN: stopped here 2021 07 20
%[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
%There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here  In one of the first empirical tests for storage effect and its requirements, \cite{Adler_2006} showed that climate variability - including extreme events - promoted the coexistence of grass species in Kansas prairie, specifically their ability to recover from low densities. Moreover, in studies on the invasibility of plants, the density and competitive balance of invasive Japanese knotweed increases as resource and thus climate variability increased.\citep{Parepa_2013} 
Environmental fluctuations are becoming more influenced by extreme climatic events (Hansen et al 2012) as extreme events increase in frequency, duration, and intensity \citep{Meehl_2004}. Extreme events (heat-waves, floods, hurricanes, etc.) alter the means and variation of environmental fluctuations, and thus have implications for coexistence and competition. Competitive output and community composition of grassland plants has been shown to be impacted by heat-waves and extreme flooding \citep{White_2000, White_2001, Hoover_2014}. Studies on the direct effect of extreme events on competing species are slim (but see above); however, the consequences of change in environmental means, which may be due to extreme events or gradual climate change in general, on coexistence is more commonly studied.  \cite{Descombes_ 2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2013} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers (citation). More variable environmental fluctuations are even less so considered compared to changes in spatial means; although, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}.

Extreme events can have important ecological impacts for controphic competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive (last longer, have higher duration) than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (similar, correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. This is a change in ATA not just from the event alone, but from the relationship between species. Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition, they do not directly consider ATA between species response caused by environment.

Interacting environmental drivers can be asymmetrically related in the extreme values (as shown in \cite{Gosh_2020_AER}). 
Environmental drivers directly influence competitive pressure experienced by populations through environmental response; competition is always a function of environmental response \citep{Chesson_2012, Ellner_2016}.
Thus we have implemented our tools for the study of ATAs to explore the importance of environmental relationships in their extremes on competition and coexistence. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we continue/move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. Moreover, the lottery model within the framework of coexistence theory provides a suitable opportunity to explore relationships in their extremes and the consequences to coexistence.

%If extreme events are more temporally extensive/expansive than regular environmental fluctuations, extreme events may cause the environmental response of two species who depend on temporally separated indicators of growth to become more correlated in the extremes. 
Extreme events can have important ecological impacts for controphic competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive (last longer, have higher duration) than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (similar, correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. This is a change in ATA not just from the event alone, but from the relationship between species. Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition, they do not directly consider ATA between species response caused by environment.
To demonstrate plausibility of ATA altering coexistence between species, we carried out long time scale simulations using a simple lottery model. 

@@

%move this paragraph to discussion
Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].

@@

@@

\begin{figure}
\caption{
%Figure displaying the noise. Three panels, left-, symmetric, and right-tail assoc noise. Use standard normal marginals.Explain in caption that alternatives for $(b_1,b_2)$ are obtained by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.Jasmin to make and insert the figure here using the includegraphics command, and then draft a caption along the lines described here.
Simple example of bivariate noise with standard normal marginals. A) is left-tail associated, B) is symmetric, and C) is right-tail associated. All three cases have the same Pearson correlation of $\rho=$, up to sampling.., yet show different symmetry of tail associaton. Asymmetrically tail associated cases are perfectly correlated in the lower (A)  or upper (C) halfs of the distribution. The variables $(b_1,b_2)$ are transformed by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.
\label{fig:noise}}
\end{figure}

\begin{figure}
\caption{%Pedagogical figure somehow displaying the different outcomes of competition in ATA and symmetric cases for an example. This will be cited from the Intro, so needs to be understandable to someone who does not know any of the ideas we introduce yet. So in particular you cannot use IGR as your metric of coexistence, you need to use something more intuitive. You will pick model parameter values and do two sims, one with ATA noise and one with symmtric noise, and show different outcomes. We should discuss and possibly take several iterations to get this figure right since it's a pedagogically important figure - if the reader is not convinced in the Intro that we are studying something that makes a difference, they are not going to slog through the Theory section, which is inevitably going to be tough going for many readers.
Populations inlfuenced by asymmetric tail association...
recover from low densities less often/ spend more time on average under one-species dominance. A) is a simulation over time of a population within a two species lottery model whose environmental response and competition is influenced by symmetric bivariate noise. B) is like A), but the population is influenced by bivariate noise with left-tailed asymmetric association. Both simulations were ran for total time of 100,000. Noticeable population density was defined at a threshold of 0.95 of total community space. Periods of noticable coexistence are marked in grey. While the population in B) had longer mean periods of coexistence, it had much longer mean periods of one-species dominace. The population in B) also recovered from low densities less often and thus has a lower ability to recover when rare and thus that community has lower coexistence.     
\label{fig:pedagog}}
\end{figure}

>>>>

%Put other possible titles below, commented, and we will select the best one at the end
%Relationships between extreme events and their influence on competition and coexistence
%Coexistence and consequences of extreme environmental relationships
%Coexistence and the consequences of extreme environmental relationships
%Coexistence and consequences of environmental relationships in their extremes
%Coexistence and implications of asymmetric environmental relationships
%Extreme [climatic] events and their implications for biodiversity
%Deductions to coexistence, extreme events, and asymmetric relationships
%Extreme relationships (relationships in their extremes) and their deductions ([negative] contributions) to coexistence
%Coexistence and relationships between environmental variables in their extremes

@@

%Old version, prior to my edits: Central to ecology is the maintenance of biodiversity in competitive ecosystems. Populations subject to strong
%competitive pressure may succumb to competitive exclusion - an idea which postulates if species within the same
%trophic level and similar niches exist, one will eventually outcompete the other (Hardin, 1960). Mediation of
%competitive pressure and exclusion allow for species to stably coexist. The criteria for coexistence and competitive
%exclusion can be broken down into contributing mechanisms provided by modern coexistence theory, a framework
%which seeks to assess and quantify mechanisms that result in coexisting species and thus the maintenance of
%biodiversity (Chesson, 1994).
%Now the new, edited version, with some additional changes requested, to be made by Jasmin, please:
The maintenance of biodiversity in competitive ecosystems is a research topic at the absolute core of modern ecology (\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
%***DAN: Jasmin, add a few classic references. See the first sentence of Ellner et al 2019, read the abstracts of those papers
%(more than the abstracts if you want), and consider using them as references here.
The competitive exclusion principle, covered, for instance, in most or all introductory ecology courses and texts, states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will outcompete 
the other and drive it extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how stable species coexistence 
nevertheless occurs, and biodiversity in maintained, in real ecosystems, through stabilizing mechanisms and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with name such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to quite a few empirical systems, including annual plants \citep{Angert_2009, Hallett_2019}, perennial plants \citep{Adler_2010}, tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{}. 
%***DAN: Jasmin, fill in Y and Y with names of systems, and accompanying reference(s), 
%in a similar way to how I filled in the Angert example.
%Please also read the abstract of the Angert reference to make sure it is appropriate to cite it in this way (and so that
%you become aware of that reference). One place to look for more references to put in for Y and Z is the top of the 
%right column on the first page of the Barabas review (Ecol Monogr, 2018).
even though it is mathematically complex in the eyes
of most ecologists (but see \cite{Ellner_2016,Ellner_2019} for efforts to make it more readily applied).
%***DAN: Jasmin, note the differences carefully, please, between your version (commented, above) and the edits I've 
%made. It's the same in main structure, making all the same points. It's just a bit more fleshed out. One of the purposes
%of an Intro is to link the reader to the appropriate background lit. You need a fair number of references and a little
%more detail than you previously gave to do that.

@@

%***DAN: stopped here 2021 07 20
%***DAN: Dan won't edit this para yet, Jasmin and I discussed and she will take another crack at it.
[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature. 
Interacting environmental drivers can be asymmetrically related in the extreme values (as shown in \cite{Gosh_2020_AER}). 
Environmental drivers directly influence competitive pressure experienced by populations through environmental response; competition is always a function of environmental response \citep{Chesson_2012, Ellner_2016}.
Thus we have implemented our tools for the study of ATAs to explore the importance of environmental relationships in their extremes on competition and coexistence. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we continue/move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. Moreover, the lottery model within the framework of coexistence theory provides a suitable opportunity to explore relationships in their extremes and the consequences to coexistence.

@@

%***DAN: I think we ought to put this in the Discussion
Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
However, the description of storage effects as the covariance between drivers is a (potentially dangerous/irresponsible?) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].

@@

Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.

@@

\begin{figure}
\caption{Figure displaying the noise. Three panels, left-, symmetric, and right-tail assoc noise. Use standard normal marginals.
Explain in caption that alternatives for $(b_1,b_2)$ are obtained by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.
Jasmin to make and insert the figure here using the includegraphics command, and then draft a caption along the lines described here.\label{fig:noise}}
\end{figure}

\begin{figure}
\caption{Pedagogical figure somehow displaying the different outcomes of competition in ATA and symmetric cases for an example.
This will be cited from the Intro, so needs to be understandable to someone who does not know any of the ideas we introduce yet.
So in particular you cannot use IGR as your metric of coexistence, you need to use something more intuitive. You will pick
model parameter values and do two sims, one with ATA noise and one with symmtric noise, and show different outcomes. We should
discuss and possibly take several iterations to get this figure right since it's a pedagogically important figure - if the reader
is not convinced in the Intro that we are studying something that makes a difference, they are not going to slog through
the Theory section, which is inevitably going to be tough going for many readers.\label{fig:pedagog}}
\end{figure}
%possible examples:
%population simulations - shows symmetric with faster fluctatuations to extremes of coexistence while asymmetric has longer periods of noticeable coexistence and dominace/exclusion; may be unclear
